Agreement Attraction in Turkish: Effects of Nominal and Verbal Plural Morphemes

### 2019-08-29 Paper/paper_draft.rmd
Updates.

<!--
Introduction:
 - Review of the key attraction findings: 
      * English. 
      * German: @ReifegersteEtAl:2016
      * Definitely Russian (Sliosar on gender). [@SlioussarMalko:2016]
      * Arabic: @TuckerEtAl:2015
      * Spanish: @LagoEtAl:2015
      * Turkish: @LagoEtAl:2018
      * Hammerly 2018 [@HammerlyEtAl:2018]
      * @LinzenLeonard:2018
      * @SmithFranck:2018
      * @ParkerPhillips:2017
      * @VasishthEtAl:2017 **
      * @ReifegersteEtAl:2016
      * @KwonSturt:2016
      * @RisticMolinaroMancini:2016
      * @SlevcMartin:2016
      * @PatsonHusband:2015
      * @Enochson:2015 **
      * @FranckColonnaRizzi:2015
      * @WagersEtAl:2009
      * Slovak: @BadeckerKuminiak:2007
      * @BockEtAl:2001
      * @FiederEtAl:2014 **
      * @Staub:2009
      * @TannerNicolBrehm:2014
      * @XiangEtAl:2013
      * @NicenboimEtAl:2018

@@

<!-- to-do: mention the Lago et al study briefly since we'll be referring to it. -->
In this paper, we report the results of two experiments about agreement attraction: The first experiment adressed a potential confound in the @Lago2018 study. In the second experiment we tested an explanation of agreement attraction in Turkish as a task-specific strategy: we tested  whether verbs with plural agreement can serve as as potential attractors.
The results show that \ldots 
<!--
pvl: no idea what the text below means
while people do indeed care more than just "matching" the forms of plural verbal. either comprehenders have some sort of restriction to not use cues from verbs upon accessing the number of the plural morpheme or there is a syntactic reason for the number feature of the plural morpheme to not "percolate." 
-->

@@

<!-- to-do: I suggest that you use multiple author names or XxxEtAl in your citation keys. That way, you won't get the number agreement wrong when writing about a paper authored by several people. -->
@Badecker2007 suggest that a process where the subject is retrieved from the memory plays a role in agreement attraction effects as well.
+
<!-- to-do: Don't use "see also"/"cf." without a clear reason. 
"Cf. X for a similar proposal" is fine, but
"cf. X" isn't helpful.

[see also @Wagers2009]. 
-->
+
<!-- to-do:
Genuine question: What makes you think that @Gordon2001 is a "cue-based retrieval mechanism" paper??
-->
+
Building on the work on cue-based parsing [@Lewis2005; @Lewis2006 among others], they argued that when readers reach the verb, they try to retrieve the subject using certain features (cues) that essentially marks the *subjecthood*. Thus, any element that matches (or partially matches) with the cues in use may cause agreement attraction regardless of their position or syntactic relation to the subject. Going back to \autoref{E2}, it is expected, then, for an object which shares features like being preverbal, being a noun phrase, ability to bear agreement with a subject to trigger agreement attraction.
<!-- to-do: Plz revise the entire paragraph above for clarity. I don't think that you're sure what you mean here. -->

<!-- to-do: Why "another" below? Did you mention another problem. -->
Another challenge for the Marking and Morphing model was the grammaticality asymmetry. Due to the spreading activation formula 
<!-- to-do: This is the first time you mention this formula, and the first time you mention that marking an morphing has anything to do with spreading activation. Please expand your explanation of M&M above.  -->
utilized in the Marking and Morphing model [see @Dell1986], this model overgeneralizes the attraction effects to both ungrammatical and grammatical sentences and predicts both \autoref{E1} and \autoref{E3} to be present empirically, which is not the case [@Wagers2009].

@@

<!-- to-do: Move this paragraph up, before you mention the challenge for "Marking and morphing". Then conclude the section with the problem to M&M. This will improve the flow of the paragraph. -->
On the other hand, cue-based retrieval models successfully predict the grammaticality asymmetry as shown by @Dillon2013 in an ACT-R simulation as well as numerous studies in self-paced reading, eye-tracking, and ERP studies [@Wagers2009; @Tucker2010; @Tanner2014; @Lago2015 among others]. This success was due to their replacement of source of the attraction. According to cue-based retrieval models, the probe, verb, present certain cues to retrieve an NP with matching features in agreement. When features of the grammatical controller of the verb do not *fully* match with the cues, no single NP will be satisfying the *agreement*. Thus, another NP in the clause can match with cues presented by the VP, such as its case, syntactic position, grammatical number. Thus, if there is no *retrieval cue* in the first place, i.e. the verb is not marked plural, there will be no need for partial match.

@@

Agreement attraction in sentence comprehension has beeen attested in 
English[@WagersEtAl:2009, @PatsonHusband:2015, @EnochsonCulbertson:2015, @HammerlyEtAl:2018]  (... more ...), 
German [@ReifegersteEtAl:2016], 
Arabic [@TuckerEtAl:2015], 
Spanish [@LagoEtAl:2015], 
Turkish [@LagoEtAl:2018], 
Slovak [@BadeckerKuminiak:2007],
Russian [@SlioussarMalko:2016], 
Korean [@KwonSturt:2016]

<!--
production:
  * @SlevcMartin:2016
theoretical:
  * @LinzenLeonard:2018
  * @SmithFranck:2018
  * @ParkerPhillips:2017
  * @VasishthEtAl:2017 **
other:
  * @RisticMolinaroMancini:2016
  * TannerBulkes:2015
  * FranckColonnaRizzi:2015
-->

Recently, @Lago2018 demonstrated agreement attraction in Turkish. They used sentences like \ref{Lagoitem}, \ldots

@Lago2018 makes use of genitive-possessive structures in the subject position, in which the possessive-marked noun is the head of the noun phrase which acts as the grammatical agreement controller, and the genitive noun serves as a potential attractor as in \autoref{Lagoitem}. In a speeded acceptability judgment study, @Lago2018 found a significant effect of number agreement attraction. However, the interpretation of their findings may be a result of the fact that non-subjecthood cues originate from their use of morphologically ambiguous forms of the possessive. In the possessive forms that are used, all the head nouns in their stimuli are ambiguous between possessive and accusative.

\begin{exe}
\ex \label{Lagoitem}
\gll \c{S}ark{\i}c{\i}-n{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}-lar.\\
singer-\textsc{Gen} vocalist-\textsc{Poss} stage-\textsc{Loc} non-stop jump-\textsc{Pst}-\textsc{3Pl}\\
\glt The singer's backup vocalist jumped on the stage non-stop.
\end{exe}

In Turkish, accusative number agreement controllers are extremely rare, while genitive agreement controllers are very frequent. Thus, @Lago2018's finding could potentially be explained with occasional shallow processing. When all syntactic relations in the sentence were processed fully, the possessive noun may have been identified as the controller. Meanwhile, the genitive noun may sometimes have been erroneously identified as the controller during shallow processing because genitives are more likely to act as agreement controllers than accusatives. In order to check this effect, I have replicated @Lago2018 in Experiment 1. I have used different head nouns which ends with a vowel to differentiate between the accusative case and the possessive suffix, i.e. *taksici* (taxi driver) instead of *vokalist* in \autoref{Lagoitem}.

<!-- -->
First of all, I investigate the effects of ambiguous suffix with Experiment 1, which is a replication of @Lago2018.

@@

In this study, I examine the question of whether agreement attraction is due to the shallow parsing that mainly depends on the memory of plural morpheme *-lAr* and matching the verbal plural morpheme with the nominal plural morpheme.

...

With regards to percolation theories, this replication should yield no difference from @Lago2018. The syntactic depth has not changed, the number information of the head and distractor have not changed. Slight differences may be expected since I have changed some items which can be found in Appendix 1, and this situation can create erroneous number information to be used in the marking process.

As for cue-based retrieval models, I should expect a lower agreement attraction effect in this replication. This prediction stems from the fact that the possibility of erroneous representation of features of the head noun is now lower due to  my efforts to differentiate between the accusative case and the possessive marking. @Lago2018 argues that due to the subjecthood information provided by the genitive case, there is an effect of agreement attraction. I eliminate the possibility of possible non-subjecthood clues given by the form that is shared by the accusative case and the possessive suffix.

\begin{exe}
  \ex \label{ouritem1}
      \gll D\"{o}v-d\"{u}k-(\textbf{ler})-i \c{c}ocuk mutfak-ta bayg{\i}n hal-de yat-{\i}yor-(\textbf{lar})-d{\i}.\\
  beat-\textsc{Nmlz}-(\textsc{Pl})-\textsc{Poss} kid kitchen-\textsc{Loc} unconscious state-\textsc{Loc} lie-\textsc{Prog}-(\textsc{Pl})-\textsc{Pst}\\
      \glt Intended: The kid that they beat were laying in the kitchen unconscious.
\end{exe}

As for cue-based retrieval models, I would expect an agreement attraction quite similar to the one that I originally have in @Lago2018. The reason for this expected agreement attraction is the fact that apart from having a verbal root, the relative clauses and nominals behave exactly the same in terms of *nounness*. Like nominals, nominalized structures first marked with plural and then with the case or possessive suffix.  From this perspective, it seems like there should be no difference. However, one may argue that the features related to the verbal and nominal can be encoded differently. Thus, there will no partial match at any time given due to the fact that verbal plural morpheme at the main verb does specify cues for a specific type of plural marking. One other question that may arise from this structure is that whether or not the parts of already-formed dependencies can feed the following dependency resolution. In other words, the verbal plural morpheme *-lar* on *dövdükleri* in \autoref{ouritem1} is already a trigger of a cue retrieval process. And, when it is merged with the nominal head *çocuk*, there is a formed semantic meaning that is  "a group of people has beaten a kid."  However, this *suggestion* assumes that relative clauses are formed at the very moment. Moreover, I also need to assume that after forming a relative clause they are unreachable, meaning that they do not contribute to the processing of the sentence, anymore.

Lastly, in  my extremely shallow processing model, I would expect a significant effect of agreement attraction. Due to one of the writers' inability to *run the numbers* and *form the mpt trees in time*, this part---unfortunately---does not go beyond than a verbal argumentation. The reason I expect a significant effect of the agreement attraction is due to the fact that upon the agreement process is driven by the faulty memory of the place of the plural morpheme and uneducated guessing. Since the parsing is shallow, comprehenders do not engage in perfect attribution of the features, thus having a somewhat blurry representation behind. I expect this behavior to be omnipresent in other agreement attraction phenomenon that has been shown, as well as in @Lago2018 and in  my replication of it. Therefore, there is no reason to have anything to be different in experiment 2.

## Experiment 1

@@

In their study, @Lago2018 investigate the comprehension of subject-verb agreement  Turkish monolinguals. They use speeded acceptability judgments for the effects of number attraction in Turkish. Their sentences make us of genitive-possessive constructions in the subject position, where the genitive is the attractor and the possessive is the head noun. They manipulate the grammaticality of the sentence by changing the plural morphology of the verb, and they also manipulate the plurality of the attractor noun. In grammatical conditions, the subject and the verb both bear singular morphology with no overt morpheme. Moreover, in the ungrammatical conditions, the verb bears the overt *-lAr* morpheme whereas the subject is still singular as exemplified below.

@@

\ex \textsc{ungrammatical, sg attractor} \label{lago4}
\gll \c{S}ark{\i}c{\i}-n{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}-lar.\\
singer-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3Pl}\\
\glt The singer's backup vocalist jumped on the stage non-stop.

\ex \textsc{grammatical, pl attractor} \label{lago2}
\gll \c{S}ark{\i}c{\i}-lar-{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}\\
singer-\textsc{pl}-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-$\varnothing$\\
\glt The singers' backup vocalist jumped on the stage non-stop.

\ex \textsc{ungrammatical, sg attractor} \label{lago3}
\gll \c{S}ark{\i}c{\i}-lar-{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}-lar.\\
singer-\textsc{pl}-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3Pl}\\
\glt The singers' backup vocalist jumped on the stage non-stop.

\ex \textsc{grammatical, pl attractor} \label{lago1}
\gll \c{S}ark{\i}c{\i}-n{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}\\
singer-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-$\varnothing$\\
\glt The singer's backup vocalist jumped on the stage non-stop.

@@

They found a significant effect of number attraction in Turkish ranging between 11\%--15\% across monolinguals. @Lago2018 stipulate that the Turkish genitive case does not provide a strong cue against subjecthood since Turkish frequently makes use of genitive marked subjects in embedded clauses, as in \autoref{gensubj}. This is in contrast with English, in which the genitive case is compatible with subjecthood. Thus, @Lago2018 argue that this robust agreement attraction effect has been linked to the case information carried by accusative and genitive.

\begin{exe}
\ex \label{gensubj}
\gll k\"{o}y-\"{u} bir haydut-un bas-t{\i}\u{g}-{\i}n-{\i} duy-du-m.\\
village-\textsc{acc} a bandit-\textsc{gen} raid-\textsc{nmlz}-\textsc{3Sg}-\textsc{acc} hear-\textsc{pst}-\textsc{1Sg}\\
\glt I heard that a (certain) robber raided the village. (Adapted from Woolford (2009))
\end{exe}

However,  my initial hypothesis postulates that participants engage in shallow processing, which results in a situation in which not only the genitive case but also the possessive case plays a significant role in delivering the case information. None of the experimental items in @Lago2018 have a head noun which ends with a vowel; thus, all of the possessive markers are morphologically and phonologically ambiguous between the accusative case and the possessive. Unlike the genitive case, it is extremely rare for the accusative case to appear on the head noun of the subject, which only occurs in raising predicates as in \autoref{dsm}.

\begin{exe}
\ex \label{dsm}
\gll Ben sen-i git-ti-n san-d{\i}-m\\
I you-\textsc{acc} go-\textsc{pst}-\textsc{2Sg} suppose-\textsc{pst}-\textsc{1Sg}\\
\glt I thought you were gone.
\end{exe}

In order to test this hypothesis, I have replicated the experiment of @Lago2018 with modified items, in which the possessive is not ambiguous with the accusative case. I tried to be as faithful as possible to the original sentences while also trying to make the sentences as plausible as possible. I keep the semantic relation between the head noun and the controller the same with what has been described in @Lago2018, which is either a relation regarding profession or a service that is given by the head noun for the possessor. A set of sentences are exemplified below.

\begin{exe}
\ex
  \begin{xlist}
  \ex \underline{ungrammatical, pl attractor} \label{exp3}
      \gll Komedyen-ler-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}-ler.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}--\textsc{pl}\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \underline{grammatical, pl attractor} \label{exp2}
      \gll Komedyen-ler-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \underline{ungrammatical, sg attractor} \label{exp4}
      \gll Komedyen-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}-ler.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-\textsc{pl}\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

  \ex \underline{grammatical, sg attractor} \label{exp1}
      \gll Komedyen-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

As seen in the examples, unlike @Lago2018's experimental items, all of  my items bear the *-sI* possessive marker instead of the ambiguous *-I* marker. As for the filler items, I could not use or modify the original experiment items since the fillers of the original study were not online. I have used two types of filler sentences: grammatical sentences in which the verb bears plural agreement (\autoref{fillera}), and ungrammatical sentences in which the verb does not bear plural agreement (\autoref{fillerb}). I wanted to nullify a possible strategy by the participants where they associate the sentence-final morpheme directly with the acceptability of the sentence. And I also wanted to eliminate the possibility of participants disregarding other elements in the sentences while answering the questions I ask right after the sentence.

\begin{exe}
\ex
\begin{xlist}
\ex \label{fillera}
\gll Adam-{\i}n anne-si fena-la\c{s}-{\i}nca inek kurban et-ti-ler.\\
man-\textsc{gen} mother-\textsc{poss} bad-\textsc{vrb}-\textsc{cvb} cow sacrifice do-\textsc{pst}-\textsc{pl}\\
\glt When his mother got ill, (they) sacrificed a cow.
\ex \label{fillerb}
\gll *Pizzac{\i}-n{\i}n kurye-si t\"{o}kezle-yince sos-lar yer-e sa\c{c}-t{\i}.\\
pizzaria-\textsc{gen} courier-\textsc{poss} trip-\textsc{cvb} sauce-\textsc{pl} floor-\textsc{dat} scatter-\textsc{pst}\\
\glt Intended: When the pizza boy tripped, sauces scattered around. 
\end{xlist}
\end{exe}

@@

One hundred and seven Turkish speakers were recruited from Bogazici University in İstanbul. We verified that Turkish was their native language and that they predominantly used it in their daily lives. 
In the experiments, participants were asked to judge the acceptability of Turkish sentences. All of the sentences were presented one word at a time in the center of the screen for 500 ms per word.
<!-- unlike @Lago2018 and @Wagers2009, in which the duration was 300 ms per word. -->
The experiment was run on *Ibex Farm*, a web-based experiment platform (\url{http://spellout.net/ibexfarm}). All experimental stimuli, data, and documentation can be found on \url{https://github.com/utkuturk/AgreementAttraction_Experiment}.

@@

In order to find out whether the noun class, and thus the presence of an ambiguity between genitive and accusative affected the magnitude of the agreement attraction effect, we analyzed the data from our experiment
together with the data from @Lago2018.

@@

Figure \@ref(fig:exp1AvgResponse) shows the percentage of "acceptable" answers as a function of sentence grammaticality and 
with the grammaticality on the x-axis, and color marking the grammatical number of the attractor noun phrase. As seen in the figure, `r round(100*avg_exp[3,6])`% of the sentences with plural attractors and a singular verb were accepted by the participants, in line with the findings of @Lago2018. I also see a number attraction rate of `r round(100*(avg_exp[3,6]-avg_exp[4,6]))`% which was also observed in @Lago2018 with close results.

@@

In Figure \@ref(fig:exp1AvgRTs), I see that average response times are significantly higher than the those of @Lago2018. I also see it in \@ref(exp1ResponseTimeModelPlot), since **b_cEndsinVowel** indicates whether or not it is Experiment 1 or Experiment from @Lago2018, i.e. distractors in items in @Lago2018 ends with a consonant, while ours ends in a vowel. Again, these results show the effect of disambiguation of the possessive suffix from the accusative case has surfaced as the difficulty in response and memory retrieval time.

@@

Experiment 1 found number attraction effects, like also attested in @Lago2018, and also found a significant effect of form ambiguity of possessive and an interaction of plural attractor with ungrammaticality in response times. Illusionary agreement of the verb with the plural attractor was more than the singular attractor. In addition to what has been discussed by @Avetisyan2019. this results also show that not only the case itself but also the form of the case can affect the response times.

@@

However, the existence of an extremely close degree of agreement attraction effects and non-significant nature of the effect of the vowel ending indicates that I cannot readily discard other explanations. Another reason for not jumping to the conclusion with percolation models can be related to the number of participants @Lago2018 have included in their study [see @Vasishth2018].

>>>>

<!--
Introduction:
 - Review of the key attraction findings: 
      * English. 
      * German: @ReifegersteEtAl:2016
      * Definitely Russian (Sliosar on gender). [@SlioussarMalko:2016]
      * Arabic: @TuckerEtAl:2015
      * Spanish: @LagoEtAl:2015
      * Turkish: @LagoEtAl:2019
      * Hammerly 2019 [@HammerlyEtAl:2019]
      * @LinzenLeonard:2018
      * @SmithFranck:2018
      * @ParkerPhillips:2017
      * @VasishthEtAl:2017 **
      * @ReifegersteEtAl:2016
      * @KwonSturt:2016
      * @RisticMolinaroMancini:2016
      * @SlevcMartin:2016
      * @PatsonHusband:2015
      * @Enochson:2015 **
      * @FranckColonnaRizzi:2015
      * @WagersEtAl:2009
      * Slovak: @BadeckerKuminiak:2007
      * @BockEtAl:2001
      * @FiederEtAl:2014 **
      * @Staub:2009
      * @TannerNicolBrehm:2014
      * @XiangEtAl:2013
      * @NicenboimEtAl:2018

@@

<!-- to-do: mention the Lago et al study briefly since we'll be referring to it. -->
In this paper, we report the results of two experiments about agreement attraction: The first experiment adressed a potential confound in the @LagoEtAl:2019 study. In the second experiment we tested an explanation of agreement attraction in Turkish as a task-specific strategy: we tested  whether verbs with plural agreement can serve as as potential attractors.
The results show that \ldots 
<!--
pvl: no idea what the text below means
while people do indeed care more than just "matching" the forms of plural verbal. either comprehenders have some sort of restriction to not use cues from verbs upon accessing the number of the plural morpheme or there is a syntactic reason for the number feature of the plural morpheme to not "percolate." 
-->

@@

<!-- to-do: I suggest that you use multiple author names or XxxEtAl in your citation keys. That way, you won't get the number agreement wrong when writing about a paper authored by several people. -->
@BadeckerKuminiak:2007 suggest that a process where the subject is retrieved from the memory plays a role in agreement attraction effects as well.

<!-- to-do: Don't use "see also"/"cf." without a clear reason. 
"Cf. X for a similar proposal" is fine, but
"cf. X" isn't helpful.

[see also @WagersEtAl:2009]. 
-->

<!-- to-do:
Genuine question: What makes you think that @Gordon2001 is a "cue-based retrieval mechanism" paper??
-->

Building on the work on cue-based parsing [@Lewis2005; @Lewis2006 among others], they argued that when readers reach the verb, they try to retrieve the subject using certain features (cues) that essentially marks the *subjecthood*. Thus, any element that matches (or partially matches) with the cues in use may cause agreement attraction regardless of their position or syntactic relation to the subject. Going back to \autoref{E2}, it is expected, then, for an object which shares features like being preverbal, being a noun phrase, ability to bear agreement with a subject to trigger agreement attraction.
<!-- to-do: Plz revise the entire paragraph above for clarity. I don't think that you're sure what you mean here. -->

<!-- to-do: Why "another" below? Did you mention another problem. -->
Another challenge for the Marking and Morphing model was the grammaticality asymmetry. Due to the spreading activation formula 
<!-- to-do: This is the first time you mention this formula, and the first time you mention that marking an morphing has anything to do with spreading activation. Please expand your explanation of M&M above.  -->
utilized in the Marking and Morphing model [see @Dell1986], this model overgeneralizes the attraction effects to both ungrammatical and grammatical sentences and predicts both \autoref{E1} and \autoref{E3} to be present empirically, which is not the case [@WagersEtAl:2009].

@@

<!-- to-do: Move this paragraph up, before you mention the challenge for "Marking and morphing". Then conclude the section with the problem to M&M. This will improve the flow of the paragraph. -->
On the other hand, cue-based retrieval models successfully predict the grammaticality asymmetry as shown by @Dillon2013 in an ACT-R simulation as well as numerous studies in self-paced reading, eye-tracking, and ERP studies [@WagersEtAl:2009; @TuckerEtAl:2015; @TannerEtAl:2014; @LagoEtAl:2015 among others]. This success was due to their replacement of source of the attraction. According to cue-based retrieval models, the probe, verb, present certain cues to retrieve an NP with matching features in agreement. When features of the grammatical controller of the verb do not *fully* match with the cues, no single NP will be satisfying the *agreement*. Thus, another NP in the clause can match with cues presented by the VP, such as its case, syntactic position, grammatical number. Thus, if there is no *retrieval cue* in the first place, i.e. the verb is not marked plural, there will be no need for partial match.

@@

Agreement attraction in sentence comprehension has beeen attested in 
Arabic [@TuckerEtAl:2015], 
Armenian [@AvetisyanEtAl:2019],
English [@WagersEtAl:2009; @PatsonHusband:2015; @EnochsonCulbertson:2015; @HammerlyEtAl:2019]  (... more ...), 
German [@ReifegersteEtAl:2016],
Korean [@KwonSturt:2016],
Russian [@SlioussarMalko:2016], 
Slovak [@BadeckerKuminiak:2007],
Spanish [@LagoEtAl:2015], 
Turkish [@LagoEtAl:2019]

<!--
production:
  * @SlevcMartin:2016
theoretical:
  * @LinzenLeonard:2018
  * @SmithFranck:2018
  * @ParkerPhillips:2017
  * @VasishthEtAl:2017 **
  * @LewisPhillips:2015
  TannerEtAl:2014
  TannerBulkes:2015
other:
  * @RisticMolinaroMancini:2016
  * TannerBulkes:2015
  * FranckColonnaRizzi:2015
-->

<!--
note: Utku's trees can go from left to right. The document below explains how:
https://en.wikibooks.org/wiki/LaTeX/Linguistics
-->

# Agreement attraction in Turkish

Recently, @LagoEtAl:2019 demonstrated agreement attraction in Turkish. In a speeded acceptability judgement study, they presented sentences like (\ref{Lagoitem}) to participants. 
@LagoEtAl:2019 makes use of genitive-possessive structures in the subject position, in which the possessive-marked head of the complex noun phrase *'the singer's vocalist'* acts as the grammatical agreement controller, and the genitive noun serves as a potential attractor.

<!-- to-do: where did the 'backup' come from in the translation ??? -->
<!-- to-do: what is the singer's vocalist ??? does this expression make sense? -->
\begin{exe}
\ex \label{Lagoitem}
\gll *[Şarkıcı-nın \underline{vokalist-i}] sahne-de sürekli \underline{zıpla-dı-lar}.\\
singer-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3pl}\\
\glt The singer's backup vocalist jumped on the stage non-stop.
\end{exe}

@LagoEtAl:2019 found a significant effect of number agreement attraction.
They argue that genitive modifiers my cause agreement attraction in Turkish because Turkish makes frequent use of genitive subjects in embedded clauses. This would explain why genitive modifiers cause little to no agreement attraction in English 
[@NicolEtAl:2016]. <!-- to-do: read, double-check the claim above -->

However, a potential complication in interpreting these results is the fact that all head nouns in Lago et al.'s stimuli were morphologically ambiguous between possessive and accusative case.
\ldots two words of explanation why consonant-ending nouns are ambiguous \ldots

In Turkish, accusative number agreement controllers are extremely rare, while genitive agreement controllers are very frequent. Thus, it is possible that the ambiguity of the head noun may occasionally render the genitive modifier a stronger contender for being the agreement controller, resulting in agreement attraction.

In order to control for the effect of head noun ambiguity on the magnitude of agreement attraction, we replicated Lago et al.'s experiment with unambiguous head nouns. To this end, we modified Lago et al.'s stimuli by replacing the head nouns with unambiguous nouns. 
\ldots which end in a vowel to differentiate between the accusative case and the possessive suffix, i.e. *taksici* (taxi driver). 
\ldots two words on why vowel-ending nouns don't exhibit this ambiguity \ldots
For example, we changed sentence (\ref{Lagoitem}) to sentence (\ref{LagoiteModified}) instead of in \autoref{Lagoitem}.

\begin{exe}
\ex \label{LagoiteModified}
\gll *[Şarkıcı-nın \underline{XXXX}] sahne-de sürekli \underline{zıpla-dı-lar}.\\
singer-\textsc{gen} XXXX-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3pl}\\
\glt The singer's XXXX jumped on the stage non-stop.
\end{exe}

<!-- to-do: 
Some more details on the modification of nouns. Like, did we
change only head nouns, or other parts too? How often?
Did we change anything except the possessive and genitive nouns? If so, how often?
-->

@@

The aim of this study was examine the question of whether agreement attraction is due to the shallow parsing that mainly depends on the memory of plural morpheme *-lAr* and matching the verbal plural morpheme with the nominal plural morpheme.

@@

In order to test this hypothesis, I have replicated the experiment of @LagoEtAl:2019 with modified items, in which the possessive is not ambiguous with the accusative case. I tried to be as faithful as possible to the original sentences while also trying to make the sentences as plausible as possible. I keep the semantic relation between the head noun and the controller the same with what has been described in @LagoEtAl:2019, which is either a relation regarding profession or a service that is given by the head noun for the possessor. A set of sentences are exemplified below.

@@

  \ex \textsc{ungrammatical, pl attractor} \label{exp1_a}
      \gll Komedyen-ler-in yardımcı-sı poyraz-dan dolayi ü\c{s}ü-dü-ler.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}--\textsc{pl}\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \textsc{grammatical, pl attractor} \label{exp1_b}
      \gll Komedyen-ler-in yardımcı-sı poyraz-dan dolayi ü\c{s}ü-dü.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \textsc{ungrammatical, sg attractor} \label{exp1_c}
      \gll Komedyen-in yardımcı-sı poyraz-dan dolayi ü\c{s}ü-dü-ler.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-\textsc{pl}\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

  \ex \textsc{grammatical, sg attractor} \label{exp1_d}
      \gll Komedyen-in yardımcı-sı poyraz-dan dolayi ü\c{s}ü-dü.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

@@

As seen in the examples, unlike @LagoEtAl:2019's experimental items, all of  my items bear the *-sI* possessive marker instead of the ambiguous *-I* marker. As for the filler items, I could not use or modify the original experiment items since the fillers of the original study were not online. I have used two types of filler sentences: grammatical sentences in which the verb bears plural agreement (\autoref{fillera}), and ungrammatical sentences in which the verb does not bear plural agreement (\autoref{fillerb}). I wanted to nullify a possible strategy by the participants where they associate the sentence-final morpheme directly with the acceptability of the sentence. And I also wanted to eliminate the possibility of participants disregarding other elements in the sentences while answering the questions I ask right after the sentence.

\begin{exe}
\ex
\begin{xlist}
\ex \label{fillera}
\gll Adam-ın anne-si fena-la\c{s}-ınca inek kurban et-ti-ler.\\
man-\textsc{gen} mother-\textsc{poss} bad-\textsc{vrb}-\textsc{cvb} cow sacrifice do-\textsc{pst}-\textsc{pl}\\
\glt When his mother got ill, (they) sacrificed a cow.
\ex \label{fillerb}
\gll *Pizzacı-nın kurye-si tökezle-yince sos-lar yer-e saç-tı.\\
pizzaria-\textsc{gen} courier-\textsc{poss} trip-\textsc{cvb} sauce-\textsc{pl} floor-\textsc{dat} scatter-\textsc{pst}\\
\glt Intended: When the pizza boy tripped, sauces scattered around. 
\end{xlist}
\end{exe}

@@

One hundred and seven Turkish speakers were recruited from Bogazici University in İstanbul. We verified that Turkish was their native language and that they predominantly used it in their daily lives. 
In the experiments, participants were asked to judge the acceptability of Turkish sentences. All of the sentences were presented one word at a time in the center of the screen for 500 ms per word.
<!-- unlike @LagoEtAl:2019 and @WagersEtAl:2009, in which the duration was 300 ms per word. -->
The experiment was run on *Ibex Farm*, a web-based experiment platform (\url{http://spellout.net/ibexfarm}). All experimental stimuli, data, and documentation can be found on \url{https://github.com/utkuturk/AgreementAttraction_Experiment}.

@@

In order to find out whether the noun class, and thus the presence of an ambiguity between genitive and accusative affected the magnitude of the agreement attraction effect, we analyzed the data from our experiment
together with the data from @LagoEtAl:2019.

@@

Figure \@ref(fig:exp1AvgResponse) shows the percentage of "acceptable" answers as a function of sentence grammaticality and 
with the grammaticality on the x-axis, and color marking the grammatical number of the attractor noun phrase. As seen in the figure, `r round(100*avg_exp[3,6])`% of the sentences with plural attractors and a singular verb were accepted by the participants, in line with the findings of @LagoEtAl:2019. I also see a number attraction rate of `r round(100*(avg_exp[3,6]-avg_exp[4,6]))`% which was also observed in @LagoEtAl:2019 with close results.

@@

In Figure \@ref(fig:exp1AvgRTs), I see that average response times are significantly higher than the those of @LagoEtAl:2019. I also see it in \@ref(exp1ResponseTimeModelPlot), since **b_cEndsinVowel** indicates whether or not it is Experiment 1 or Experiment from @LagoEtAl:2019, i.e. distractors in items in @LagoEtAl:2019 ends with a consonant, while ours ends in a vowel. Again, these results show the effect of disambiguation of the possessive suffix from the accusative case has surfaced as the difficulty in response and memory retrieval time.

@@

Experiment 1 found number attraction effects, like also attested in @LagoEtAl:2019, and also found a significant effect of form ambiguity of possessive and an interaction of plural attractor with ungrammaticality in response times. Illusionary agreement of the verb with the plural attractor was more than the singular attractor. In addition to what has been discussed by @AvetisyanEtAl:2019. this results also show that not only the case itself but also the form of the case can affect the response times.

@@

However, the existence of an extremely close degree of agreement attraction effects and non-significant nature of the effect of the vowel ending indicates that I cannot readily discard other explanations. Another reason for not jumping to the conclusion with percolation models can be related to the number of participants @LagoEtAl:2019 have included in their study [see @Vasishth2018].

---

\begin{exe}
  \ex \label{ouritem1}
      \gll Döv-dük-(\textbf{ler})-i çocuk mutfak-ta baygın hal-de yat-ıyor-(\textbf{lar})-dı.\\
  beat-\textsc{Nmlz}-(\textsc{Pl})-\textsc{Poss} kid kitchen-\textsc{Loc} unconscious state-\textsc{Loc} lie-\textsc{Prog}-(\textsc{Pl})-\textsc{Pst}\\
      \glt Intended: The kid that they beat were laying in the kitchen unconscious.
\end{exe}
